---
title: "NOAAnalysis"
author: "Katalin Virag"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{NOAAnalysis}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, echo = FALSE, message = FALSE}
library(NOAAnalysis)
library(magrittr)
library(chron)
g <- system.file("extdata", package = "NOAAnalysis")
knitr::opts_knit$set(root.dir = g)
```

These functions can be used to work with the NOAA Significant Earthquakes dataset from [link](https://www.ngdc.noaa.gov/nndc/struts/form?t=101650&s=1&d=1).

## Functions included in the package

* `eq_clean_data()` to clean data
* `eq_location_clean()` to clean the `LOCATION_NAME` column of the NOAA data
* `geom_timeline()` to visualize the times at which earthquakes occur
* `geom_timeline_label()` to add annotation to the timeline plot of earthquakes
* `theme_timeline` to add simple theme to the timeline plot of earthquakes
* `eq_map()` to visualize the earthquakes in space
* `eq_create_label()` to create an HTML label to annotate the leaflet map

## Data

This package includes one dataset called `signif.txt` which contains information on destructive earthquakes (see [link](https://www.ngdc.noaa.gov/nndc/struts/form?t=101650&s=1&d=1)).

## Data cleaning functions

### Clean the NOAA Significant Earthquake Database with `eq_clean_data()`

This function takes raw NOAA data frame and returns a clean data frame with the following features:

* A `DATE` column created by uniting the year, month, day and converting it to the `Date` class

* `LATITUDE` and `LONGITUDE` columns converted to `numeric` class

* `LOCATION_NAME_cleaned` column from `LOCATION_NAME` by stripping out the country name and in title case

```{r message = FALSE}
noaa <- readr::read_delim(system.file("extdata", "signif.txt",
        package = "NOAAnalysis"), delim = "\t") %>% eq_clean_data()
head(noaa)
summary(noaa)
```

### Clean the LOCATION_NAME string with `eq_location_clean()`

This helper function cleans the `LOCATION_NAME` string by stripping out the country name and converts names to title case

```{r}
eq_location_clean("SYRIA: UGARIT")
```

## Visualization tools

### Visualize the earthquakes in time with `geom_timeline()` and apply simple theme with `theme_timeline`

The `geom_timeline()` function creates a `ggplot` object which shows the times at which earthquakes occur ranging from `xmin` to `xmax` dates with a point for each earthquake. The `x` aesthetic is a date and an optional `y` aesthetic is a factor indicating some stratification (e.g. country). Additionally, the magnitudes and the number of deaths associated with each earthquake can be shown.

The `theme_timeline` creates simple theme for the timeline plot.

```{r message = FALSE, fig.height = 4, fig.width = 6, fig.align = "center"}
readr::read_delim(system.file("extdata", "signif.txt",
       package = "NOAAnalysis"), delim = "\t") %>%
eq_clean_data() %>%
dplyr::filter(COUNTRY == "USA" & YEAR >= 2000) %>%
ggplot2::ggplot(ggplot2::aes(x = DATE, y = COUNTRY,
         xmin = as.Date("2000-01-01"), xmax = as.Date("2016-01-01"),
         size = as.numeric(EQ_PRIMARY), color = as.numeric(TOTAL_DEATHS))) +
geom_timeline() +
theme_timeline +
ggplot2::labs(colour = "# deaths", size = "Richter scale value")
```

Annotations can be added to the timeline plot with the `geom_timeline_label()` function. There is an option to subset to `n_max` number of earthquakes, where we take the `n_max` largest (by magnitude) earthquakes.

```{r message = FALSE, fig.height = 4, fig.width = 6, fig.align = "center"}
readr::read_delim(system.file("extdata", "signif.txt",
       package = "NOAAnalysis"), delim = "\t") %>%
eq_clean_data() %>%
dplyr::filter(COUNTRY %in% c("USA", "CHINA") & YEAR >= 2000) %>%
ggplot2::ggplot(ggplot2::aes(x = DATE, y = COUNTRY,
                xmin = as.Date("2000-01-01"), xmax = as.Date("2016-01-01"),
                size = as.numeric(EQ_PRIMARY), color = as.numeric(TOTAL_DEATHS))) +
geom_timeline() +
geom_timeline_label(ggplot2::aes(label = LOCATION_NAME_cleaned)) +
theme_timeline +
ggplot2::labs(colour = "# deaths", size = "Richter scale value")
```

### Visualize the earthquakes in space with `eq_map()`

`eq_map()` creates a `leaflet` object which maps and annotates the epicenters. Each earthquake is shown with a circle, and the radius of the circle is proportional to the earthquake's magnitude. Each point is annotated in pop up window containing annotation data stored in the `annot_col` column of data.

```{r message = FALSE, fig.height = 4, fig.width = 6, fig.align = "center"}
readr::read_delim(system.file("extdata", "signif.txt",
       package = "NOAAnalysis"), delim = "\t") %>%
eq_clean_data() %>%
dplyr::filter(COUNTRY == "MEXICO" & YEAR >= 2000) %>%
eq_map(annot_col = "DATE")
```

More annotation can be added to the `leaflet` map with the `eq_create_label()` function. This function puts together a character string for each earthquake that showes the cleaned location, the magnitude, and the total number of deaths (if not missing).

```{r message = FALSE, fig.height = 4, fig.width = 6, fig.align = "center"}
readr::read_delim(system.file("extdata", "signif.txt",
       package = "NOAAnalysis"), delim = "\t") %>%
eq_clean_data() %>%
dplyr::filter(COUNTRY == "MEXICO" & YEAR >= 2000) %>%
dplyr::mutate(popup_text = eq_create_label(.)) %>%
eq_map(annot_col = "popup_text")
```
